---
title: "Get started with bsync"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with bsync}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bsync)
```

**bsync** provides a modern, high-efficiency toolkit for analyzing **interpersonal and behavioral
synchrony** in continuous dyadic time series. It implements windowed estimators of nonstationary
lead–lag structure — Windowed Cross-Correlation (WCC), Windowed Dynamic Time Warping (WDTW), and
Windowed Granger Causality (WGC) — together with a full preprocessing pipeline, surrogate
significance testing, optima extraction, and parameter guidance tools.

This vignette walks you through the complete workflow from raw data to a quantified leadership
index. It also gives you the decision table you need to choose the right estimator and points you
toward the relevant deep-dive articles.

## The bsync analysis workflow

A complete behavioral synchrony analysis has five stages:

1. **Preprocess** — smooth, trim, and compute kinematics from raw position data.
2. **Choose parameters** — select window size and lag ceiling that match the signal's timescales.
3. **Estimate** — run WCC, WDTW, or WGC on the prepared signals.
4. **Test significance** — compare the observed statistic against a surrogate null distribution.
5. **Quantify leadership** — extract optima and compute the Leadership Asymmetry Index.

The following example walks through all five stages using the built-in `sim_dyad` dataset, which
contains 30 seconds of simulated 3D motion-tracking data (80 Hz) for two individuals. Person A
leads a rhythmic movement early in the recording, they briefly synchronize, and Person B takes the
lead by the end.

## Step 1: Preprocess

Raw positional data almost always needs smoothing before analysis. High-frequency noise amplifies
when derivatives are calculated, contaminating the cross-correlations. We apply a Savitzky-Golay
filter and then compute the 1D velocity along the Z-axis.

```{r preprocess}
library(dplyr)

# Smooth raw positions, then compute 1D velocity
df <- sim_dyad |>
  mutate(
    z_A_sm = smooth_signal(z_A, method = "sgolay", window = 5),
    z_B_sm = smooth_signal(z_B, method = "sgolay", window = 5),
    vel_A  = calc_velocity_1d(time, z_A_sm, fill_edges = TRUE),
    vel_B  = calc_velocity_1d(time, z_B_sm, fill_edges = TRUE)
  )
```

Other preprocessing helpers available in bsync: `trim_edges()` (drop polynomial-smoothing
artifacts), `aggregate_by_time()` / `downsample_signal()` (reduce sample rate),
`diagnose_ts_gaps()` / `impute_ts_gaps()` (handle missing values), and `evaluate_signal_power()`
(PSD-based downsampling guidance; see `vignette("determine-downsampling")`).

## Step 2: Choose parameters

The two key hyperparameters for WCC — `window_size` (samples per window) and `lag_max` (maximum
temporal offset to test) — should be matched to the dominant behavioral timescale. `bsync`
provides three complementary tools (detailed in `vignette("choosing-parameters")`):

* **`suggest_wcc_params()`** — PSD-driven starting values for a single dyad.
* **`synchrony_multiverse()`** — sweep the full grid and view the specification curve.
* **`autotune_wcc()`** — select parameters that generalize across a multi-dyad dataset.

For this quick-start we use hard-coded values that match the 0.5 Hz dominant cycle in `sim_dyad`
(window ≈ 4 cycles at 80 Hz = 640 samples; lag ≈ half the window = 320 samples). In practice,
always derive these from your own signal.

## Step 3: Estimate

With the data prepared and parameters chosen, run `wcc()`. The function returns a `wcc_res`
object that carries the full windowed surface, the aggregate Fisher's Z, and all settings.

```{r estimate}
wcc_res <- wcc(
  x                = df$vel_A,
  y                = df$vel_B,
  window_size      = 640,
  lag_max          = 320,
  window_increment = 80,
  lag_increment    = 1
)

summary(wcc_res)
```

Plot the surface to see the shifting correlation landscape:

```{r plot-surface, fig.width = 7, fig.height = 4}
plot(wcc_res, time_step = 1 / 80) # time_step converts frames → seconds
```

The `plot()`, `print()`, `summary()`, `tidy()`, and `glance()` methods work identically for
`wdtw_res` and `wgranger_res` objects — the shared `bsync_surface` superclass ensures a
consistent API across all three estimators.

## Step 4: Test significance

Because behavioral time series are autocorrelated, a high cross-correlation can occur purely by
chance. Surrogate testing builds an empirical null distribution by breaking the dyadic alignment
while preserving each person's individual signal characteristics.

```{r surrogate, message = FALSE}
set.seed(2026)

# Generate 200 circular-shift surrogates (use >= 1000 for publication)
null_mat <- generate_surrogate_circular(
  y            = df$vel_B,
  n_surrogates = 200L,
  lag_max      = 320
)

# Compare observed WCC against the null
surr_res <- wcc_surrogate(
  x                = df$vel_A,
  y                = df$vel_B,
  y_surrogates     = null_mat,
  window_size      = 640,
  lag_max          = 320,
  window_increment = 80,
  lag_increment    = 1
)

print(surr_res)
```

For large or computationally expensive analyses, set a parallel plan before running:
`future::plan(future::multisession, workers = 4)`. See `vignette("surrogate-testing")` for a
thorough treatment of surrogate methods and when to use phase randomization instead.

## Step 5: Extract optima and quantify leadership

The full WCC surface tells us *how much* synchrony exists at each moment. To track *who is
leading*, we extract the lag that maximizes correlation within each window (`pick_optima()`),
then summarize those lags into a continuous Leadership Asymmetry Index (`leadership_asymmetry()`).

```{r optima, fig.width = 7, fig.height = 4}
wcc_res |>
  pick_optima(L_size = 9) |>
  leadership_asymmetry(epoch_size = 3, min_valid = 1) |>
  plot(smooth = TRUE)
```

The LAI ranges from −1 (y leads entirely) to +1 (x leads entirely). In this plot you can see
Person A leading early, a transition period in the middle, and Person B leading by the end —
exactly as simulated in `sim_dyad`.

Visualize the optima overlaid on the surface with `plot_optima_overlay()`:

```{r overlay, fig.width = 7, fig.height = 4}
optima_df <- pick_optima(wcc_res, L_size = 9)
plot_optima_overlay(wcc_res, optima_df, time_step = 1 / 80, show_zero_lag = TRUE)
```

---

## Which estimator should I use?

| | **WCC** | **WDTW** | **WGC** |
|---|---|---|---|
| **Measures** | Linear shape & amplitude similarity | Shape similarity (time-warped) | Directed predictive coupling |
| **Best for** | Continuous kinematics, vocal pitch, physiological signals with known lag range | Mimicry where one person is faster/slower | Proving direction of information flow |
| **Not for** | Variable execution speeds; categorical data | Exact-timing analyses; very noisy data | Unknown lags; non-linear systems; categorical data |
| **Output** | Correlation surface (higher = more synchrony) | Distance surface (lower = more synchrony) | F-statistic surface, both directions |
| **Key parameter** | `window_size`, `lag_max` | `window_size`, `lag_max` | `window_size`, `ar_order` |
| **Deep dive** | `vignette("wcc-workflow")` | `vignette("wdtw-workflow")` | `vignette("wgranger-workflow")` |

If you are unsure, start with WCC — it is the most widely used estimator in behavioral synchrony
research and the one for which `suggest_wcc_params()` and `autotune_wcc()` provide the most
direct guidance.

---

## Where to go next

| Article | What you will learn |
|---|---|
| `vignette("wcc-workflow")` | Full WCC pipeline with optima, LAI, tidy interface, and the two aggregate statistics (`mean_abs_z` vs. `peak`) |
| `vignette("wdtw-workflow")` | WDTW step-by-step with surrogate testing and optima extraction |
| `vignette("wgranger-workflow")` | WGC step-by-step with F-statistic and p-value plots |
| `vignette("choosing-parameters")` | Data-driven parameter selection: `suggest_wcc_params()`, `synchrony_multiverse()`, `autotune_wcc()`, and `select_specification()` |
| `vignette("surrogate-testing")` | When to use circular-shift vs. phase-randomization surrogates; best practices |
| `vignette("determine-downsampling")` | How to choose a biologically appropriate sample rate from PSD diagnostics |
