---
title: "Choosing Analysis Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Analysis Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bsync)
```

## The Parameter-Selection Problem

There is no single "correct" set of WCC hyperparameters. The optimal window size and lag ceiling
depend on the timescale of the behavior you are measuring, and those timescales vary enormously:
a mother-infant gaze interaction unfolds on a completely different timescale than a high-speed
athletic exchange.

This vignette shows three tools `bsync` provides to address this problem, in order from
quick-and-informed to thorough-and-multi-dyad:

1. **`suggest_wcc_params()`** — PSD-based starting values for a single dyad.
2. **`synchrony_multiverse()`** — sweep the full parameter grid and view the specification curve.
3. **`autotune_wcc()`** — automatically select parameters that generalize across a dyad dataset.

`select_specification()` underpins step 3 and is exported for advanced users who want to apply the
selection rule directly to a list of multiverse results.

> **Note:** Steps 1 and 3 are WCC-specific helpers. `synchrony_multiverse()` in step 2 supports
> all three estimators (`"wcc"`, `"wdtw"`, `"wgranger"`), so you can run a parameter sweep for any
> estimator.

## 1. Data-Driven Starting Values: `suggest_wcc_params()`

When you have a representative series in hand, let the signal tell you its own dominant timescale
via power spectral density (PSD):

```{r suggest}
fs <- 80 # sampling rate of sim_dyad (Hz)

params <- suggest_wcc_params(
  x           = sim_dyad$z_A,
  y           = sim_dyad$z_B,
  sample_rate = fs
)
params
```

The PSD-based cutoff captures 95% of the signal's power below ~1.25 Hz (a 0.8 s timescale). The
4-cycles-per-window heuristic (Boker et al., 2002) then gives
`window_size = round(0.8 * 4 * 80) = 256 samples`. Because the default `max_delay_sec`
(3 s = 240 samples) exceeds `window_size / 2`, the SUSY lag-cap rule caps `lag_max` at 128 samples
and announces it -- one of three hard constraints that are applied and reported, never silently
(see `?suggest_wcc_params`):

- `lag_max <= window_size / 2` (SUSY lag-cap rule)
- `window_size <= length(x) / 2` (series-length ceiling)
- Minimum-samples floor for a stable *r*

The data-driven cutoff reflects all of the signal's power, including higher-frequency content, so it
favors a shorter window. If you already know the behavioral timescale from theory -- here, the
0.5 Hz interaction cycle that `sim_dyad` was built from -- pass it directly as an override, which
yields a longer 640-sample window:

```{r suggest-override}
params_expert <- suggest_wcc_params(
  x                  = sim_dyad$z_A,
  y                  = sim_dyad$z_B,
  sample_rate        = fs,
  event_duration_sec = 2 # 0.5 Hz cycle = 2 s
)
params_expert
```

Either way, plug the result directly into `wcc()`:

```{r wcc-run}
results <- wcc(
  x                = sim_dyad$z_A,
  y                = sim_dyad$z_B,
  window_size      = params$window_size,
  lag_max          = params$lag_max,
  window_increment = params$window_increment,
  lag_increment    = params$lag_increment
)
glance(results)
```

### Window overlap (`overlap_pct`)

`suggest_wcc_params()` also derives a `window_increment` from `overlap_pct`, which controls how
far the window shifts on each step:

- **0% overlap** (`overlap_pct = 0`): the window jumps completely forward — fastest, but lowest
  temporal resolution.
- **50–75% overlap** (the default `0.5`): a balanced choice for exploratory analysis.
- **Increment of 1** (near-complete overlap): highest resolution, and feasible for typical datasets
  because `bsync` uses a C++ prefix-sum core.

## 2. Visualizing the Specification Curve: `synchrony_multiverse()`

A single parameter set is a single analytic choice. The synchrony multiverse shows you how your
conclusion changes across the full grid of defensible choices. The headline metric is **effect
size vs. the matched-null surrogate**, not raw synchrony (which autocorrelation inflates).

```{r multiverse, fig.width = 7, fig.height = 5}
# Sweep window sizes 0.5 s to 4 s and lag from 0.25 s to 1 s
set.seed(2026)
mv <- synchrony_multiverse(
  x            = sim_dyad$z_A,
  y            = sim_dyad$z_B,
  estimator    = "wcc",
  sample_rate  = 80,
  window_sec   = c(0.5, 1, 2, 4),
  lag_sec      = c(0.25, 0.5, 1),
  n_surrogates = 100L
)

mv
```

```{r multiverse-glance}
glance(mv)
```

```{r multiverse-plot, fig.width = 7, fig.height = 6}
plot(mv)
```

The specification curve (top panel) sorts parameter specifications by effect size; the choice
dashboard (bottom panel) reveals which analytic choices drive the result. If synchrony is robust,
most cells should be significant regardless of the specific hyperparameters.

The full specification grid is available as a tibble via `tidy()` or `as_tibble()`:

```{r multiverse-tidy}
head(tidy(mv))
```

### Running a multiverse for WDTW or Granger

`synchrony_multiverse()` is not WCC-only. Pass `estimator = "wdtw"` or `estimator = "wgranger"`
to sweep the same grid for those estimators. For Granger, the result carries two effect-size
columns (`es_xy` and `es_yx`, one per causal direction) and no lag axis:

```{r multiverse-granger, fig.width = 7, fig.height = 5}
set.seed(2026)
mv_g <- synchrony_multiverse(
  x            = sim_dyad$z_A,
  y            = sim_dyad$z_B,
  estimator    = "wgranger",
  sample_rate  = 80,
  window_sec   = c(1, 2, 4),
  lag_sec      = 0, # ignored for Granger; ar_order used instead
  n_surrogates = 100L
)

glance(mv_g)
```

## 3. Multi-Dyad Parameter Selection: `autotune_wcc()`

When you have a dataset of multiple dyads, `autotune_wcc()` selects parameters that are both
detectable (significant vs. the null in many dyads) and stable (consistent across dyads with
different signal characteristics).

```{r autotune}
# Simulate a dataset: three dyads sharing the same 0.5 Hz synchrony structure
set.seed(42)
dyad_list <- replicate(
  3,
  list(x = sim_dyad$z_A, y = sim_dyad$z_B),
  simplify = FALSE
)

best <- autotune_wcc(
  dyad_list    = dyad_list,
  sample_rate  = 80,
  window_sec   = c(1, 2, 4),
  lag_sec      = c(0.5, 1),
  n_surrogates = 100L
)

best
```

Plug the result directly into `wcc()`:

```{r autotune-run}
results_tuned <- wcc(
  x                = sim_dyad$z_A,
  y                = sim_dyad$z_B,
  window_size      = best$window_size,
  lag_max          = best$lag_max,
  window_increment = best$window_increment,
  lag_increment    = best$lag_increment
)
glance(results_tuned)
```

### How the selection rule works

`autotune_wcc()` applies a two-stage rule:

1. **Detectability gate.** A parameter cell must be significant (p < .05 vs. the matched-null
   surrogate) in at least `sig_pct` (default 50%) of dyads. Cells that fail this gate are
   excluded.

2. **Stability-penalized score.** Among gated cells, the score is
   `median(ES) - iqr_penalty * IQR(ES)`. This rewards high median ES but penalizes spread,
   so a parameter set that is consistent across dyads beats one that is spectacular for one dyad
   but weak for others.

If no cell passes the gate (insufficient data or too few surrogates), a warning is issued and
the highest-median-ES cell is returned as a fallback.

### Advanced: `select_specification()` directly

`select_specification()` is the exported building block that `autotune_wcc()` wraps. You can call
it directly if you have already run `synchrony_multiverse()` per dyad and want to apply the
selection rule yourself — for example, to experiment with different `sig_pct` or `iqr_penalty`
values without re-running the full sweep:

```{r select-spec}
# Run synchrony_multiverse() for each dyad
set.seed(42)
mv_list <- lapply(dyad_list, function(d) {
  synchrony_multiverse(
    x            = d$x,
    y            = d$y,
    estimator    = "wcc",
    sample_rate  = 80,
    window_sec   = c(1, 2, 4),
    lag_sec      = c(0.5, 1),
    n_surrogates = 100L
  )
})

# Apply the selection rule directly
best_adv <- select_specification(
  mv_list     = mv_list,
  sig_pct     = 0.5,
  iqr_penalty = 0.5
)

best_adv
```

### Choosing the grid to sweep

Start with a range informed by `suggest_wcc_params()` on a representative dyad. A practical
workflow:

```{r workflow}
# Step 1: get a data-driven starting point
params2 <- suggest_wcc_params(
  x           = sim_dyad$z_A,
  y           = sim_dyad$z_B,
  sample_rate = 80
)

# Step 2: build a grid around that starting point
w_base <- params2$window_size / 80 # convert to seconds
grid_windows <- unique(round(w_base * c(0.5, 1, 2), 1))
grid_lags <- unique(round(grid_windows / 2, 1))

cat("window_sec:", grid_windows, "\n")
cat("lag_sec:   ", grid_lags, "\n")
```

```{r workflow-autotune}
set.seed(42)
best2 <- autotune_wcc(
  dyad_list    = dyad_list,
  sample_rate  = 80,
  window_sec   = grid_windows,
  lag_sec      = grid_lags,
  n_surrogates = 100L
)

best2
```

## References

Boker, S. M., Xu, M., Rotondo, J. L., & King, K. (2002). Windowed cross-correlation and peak
picking for the analysis of variability in the association between behavioral time series.
*Psychological Methods*, *7*(3), 338--355.

Tschacher, W., & Meier, D. (2020). Physiological synchrony in psychotherapy sessions.
*Psychotherapy Research*, *30*(5), 558--573.

## See also

- `vignette("wcc-workflow")` for the full WCC analytical pipeline
- `vignette("surrogate-testing")` for details on circular-shift vs. phase-randomization surrogates
- `vignette("determine-downsampling")` for PSD-based guidance on choosing a sample rate
