---
title: "Choosing WCC Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing WCC Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bsync)
```

## The Parameter-Selection Problem

There is no single "correct" set of WCC hyperparameters. The optimal window size and lag ceiling depend on the timescale of the behavior you are measuring, and those timescales vary enormously:
a mother-infant gaze interaction unfolds on a completely different timescale than a high-speed athletic exchange.

This vignette shows three tools `bsync` provides to address this problem, in order from quick-and-informed to thorough-and-multi-dyad:

1. **`suggest_wcc_params()`** — PSD-based starting values for a single dyad.
2. **`synchrony_multiverse()`** — sweep the full parameter grid and see the specification curve.
3. **`autotune_wcc()`** — automatically select parameters that generalize across a dyad dataset.

## 1. Data-Driven Starting Values: `suggest_wcc_params()`

When you have a representative series in hand, let the signal tell you its own dominant timescale via power spectral density (PSD):

```{r suggest, eval = FALSE}
fs <- 80 # sampling rate of sim_dyad (Hz)

params <- suggest_wcc_params(
  x           = sim_dyad$z_A,
  y           = sim_dyad$z_B,
  sample_rate = fs
)
```

The PSD identifies the 0.5 Hz dominant cycle (~2 s period). The 4-cycles-per-window heuristic
(Boker et al., 2002) then gives `window_size = round(2 * 4 * 80) = 640 samples`, subject to the
hard constraints (see `?suggest_wcc_params`).

If you already know the behavioral timescale from theory, pass it directly:

```{r suggest-override, eval = FALSE}
params_expert <- suggest_wcc_params(
  x                  = sim_dyad$z_A,
  y                  = sim_dyad$z_B,
  sample_rate        = fs,
  event_duration_sec = 2 # 0.5 Hz cycle = 2 s
)
```

Either way, plug the result directly into `wcc()`:

```{r wcc-run, eval = FALSE}
results <- wcc(
  x                = sim_dyad$z_A,
  y                = sim_dyad$z_B,
  window_size      = params$window_size,
  lag_max          = params$lag_max,
  window_increment = params$window_increment,
  lag_increment    = params$lag_increment
)
```

## 2. Visualizing the Specification Curve: `synchrony_multiverse()`

A single parameter set is a single analytic choice. The synchrony multiverse shows you how your conclusion changes across the full grid of defensible choices. The headline metric is **effect size vs. the matched-null surrogate**, not raw synchrony (which autocorrelation inflates).

```{r multiverse, eval = FALSE}
# Sweep window sizes from 0.5 s to 4 s and lag from 0.5 s to 1 s
mv <- synchrony_multiverse(
  x            = sim_dyad$z_A,
  y            = sim_dyad$z_B,
  estimator    = "wcc",
  sample_rate  = 80,
  window_sec   = c(0.5, 1, 2, 4),
  lag_sec      = c(0.5, 1),
  n_surrogates = 100L
)

glance(mv) # robustness summary
plot(mv) # specification curve
```

The specification curve (top panel) shows effect sizes sorted by magnitude; the choice dashboard
(bottom panel) reveals which analytic choices drive the result. If synchrony is robust, most cells
should be significant regardless of the specific hyperparameters.

## 3. Multi-Dyad Parameter Selection: `autotune_wcc()`

When you have a dataset of multiple dyads, `autotune_wcc()` selects parameters that are both
detectable (significant vs. the null in many dyads) and stable (consistent across dyads with
different signal characteristics).

```{r autotune, eval = FALSE}
# Simulate a dataset: three dyads sharing the same 0.5 Hz synchrony structure
set.seed(42)
dyad_list <- replicate(
  3,
  list(x = sim_dyad$z_A, y = sim_dyad$z_B),
  simplify = FALSE
)

best <- autotune_wcc(
  dyad_list    = dyad_list,
  sample_rate  = 80,
  window_sec   = c(1, 2, 4),
  lag_sec      = c(0.5, 1),
  n_surrogates = 100L
)

# Plug the result directly into wcc()
results <- wcc(
  x                = sim_dyad$z_A,
  y                = sim_dyad$z_B,
  window_size      = best$window_size,
  lag_max          = best$lag_max,
  window_increment = best$window_increment,
  lag_increment    = best$lag_increment
)
```

### How the selection rule works

`autotune_wcc()` applies a two-stage rule:

1. **Detectability gate.** A parameter cell must be significant (p < .05 vs. the matched-null
   surrogate) in at least `sig_pct` (default 50%) of dyads. Cells that fail this gate are
   excluded.

2. **Stability-penalized score.** Among gated cells, the score is
   `median(ES) - iqr_penalty * IQR(ES)`. This rewards high median ES but penalizes spread,
   so a parameter set that is consistent across dyads beats one that is spectacular for one dyad
   but weak for others.

If no cell passes the gate (insufficient data or too few surrogates), a warning is issued and
the highest-median-ES cell is returned as a fallback.

### Choosing the grid to sweep

Start with a range informed by `suggest_wcc_params()` on a representative dyad. A practical
workflow:

```{r workflow, eval = FALSE}
# Step 1: get a data-driven starting point
params <- suggest_wcc_params(
  x           = sim_dyad$z_A,
  y           = sim_dyad$z_B,
  sample_rate = 80
)

# Step 2: build a grid around that starting point
w_base <- params$window_size / 80 # convert to seconds
grid_windows <- unique(round(w_base * c(0.5, 1, 2), 1))
grid_lags <- unique(round(grid_windows / 2, 1))

# Step 3: autotune across dyads
best <- autotune_wcc(
  dyad_list = dyad_list,
  sample_rate = 80,
  window_sec = grid_windows,
  lag_sec = grid_lags,
  n_surrogates = 200L
)
```

## References

Boker, S. M., Xu, M., Rotondo, J. L., & King, K. (2002). Windowed cross-correlation and peak
picking for the analysis of variability in the association between behavioral time series.
*Psychological Methods*, *7*(3), 338-355.

Tschacher, W., & Meier, D. (2020). Physiological synchrony in psychotherapy sessions.
*Psychotherapy Research*, *30*(5), 558-573.
