---
title: "Surrogate Testing for Significance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surrogate Testing for Significance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In interpersonal synchrony analysis, finding a high correlation or a low alignment cost is only the first step. Because behavioral time series are inherently autocorrelated, standard statistical thresholds (like $r > 0$) are often misleading. A high correlation might occur purely by chance because both signals share the same slow-moving trends.

Surrogate testing allows us to build an empirical null distribution—a "chance baseline"—to determine if our observed results are truly meaningful.

## 1. Choosing a Surrogate Method

The **bsync** package provides two primary methods for generating null data. Choosing the right one depends on the nature of your signal.

### 1.1 Circular Shift (`generate_surrogate_circular`)
The circular shift method takes a time series and shifts it by a random interval, wrapping the end of the series back to the beginning.

* **How it works:** It destroys the temporal alignment between Person A and Person B while perfectly preserving the exact amplitude distribution and the power spectrum of the original signal.
* **When to use it:** This is our default recommendation for most dyadic interaction data. It is computationally efficient and effectively breaks the cross-signal dependency while keeping the "rhythm" of the individual intact.
* **Important Note:** If your data has strong global trends (e.g., a person slowly moving across the room over 60 seconds), a simple circular shift might not break that trend. Ensure you use the `lag_max` argument to ensure shifts are large enough to decouple the signals.

### 1.2 Phase Randomization (`generate_surrogate_phase`)
Phase randomization uses the Fourier Transform to decompose a signal into its frequency components, randomizes the phases, and then reconstructs the signal.

* **How it works:** It creates a signal that has the exact same power spectrum (frequency content) as the original, but a totally randomized temporal structure.
* **When to use it:** This is the "gold standard" for highly oscillatory data (e.g., rhythmic postural sway, physiological signals like heart rate or respiration).
* **Important Note:** Because this method creates an entirely new signal from frequency components, it is statistically more conservative than circular shifting. It is significantly more computationally expensive to generate than circular shifts.

## 2. Theoretical Pipeline

Regardless of the generation method, the logic of the **bsync** surrogate pipeline remains consistent:

1. **Decouple:** Generate a matrix of $N$ null time series that remove the synchrony while keeping the individual characteristics.
2. **Evaluate:** Run your chosen synchrony metric (WCC, WDTW, or WGC) against every column in that matrix.
3. **Compare:** Calculate the empirical p-value as the proportion of surrogate results that are as extreme as, or more extreme than, your observed result.

```{r example, eval = FALSE}
library(bsync)

# 1. Generate 1000 surrogates using your preferred method
null_matrix <- generate_surrogate_circular(
  y = dyad_data$person_B,
  n_surrogates = 1000
)

# 2. Compare observed results to the null distribution
results <- wcc_surrogate(
  x = dyad_data$person_A,
  y = dyad_data$person_B,
  y_surrogates = null_matrix,
  window_size = 90,
  lag_max = 45
)

# 3. Interpret the empirical p-value
print(results)
```

## 3. Best Practices

* **Permutation Count:** For a stable p-value, we recommend at least 1,000 permutations. Using fewer (e.g., 100) is fine for rapid exploratory analysis, but it will lack the granularity needed to report significance levels like $p < .001$.
* **Power and Speed:** Surrogate analysis is "embarrassingly parallel." For heavy analyses like WDTW, always use the `future` package to distribute these 1,000+ calculations across your machine's CPU cores.
* **Interpretation:** A significant p-value does not mean the behavior is *causal*; it only means the observed degree of synchrony is statistically unlikely to have occurred if the two participants were moving independently.
