---
title: "Surrogate Testing for Significance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surrogate Testing for Significance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In interpersonal synchrony analysis, finding a high correlation or a low alignment cost is only the first step. Because behavioral time series are inherently autocorrelated, standard statistical thresholds (like $r > 0$) are often misleading. A high correlation might occur purely by chance because both signals share the same slow-moving trends.

Surrogate testing allows us to build an empirical null distribution—a "chance baseline"—to determine if our observed results are truly meaningful.

## 1. Choosing a Surrogate Method

The **bsync** package provides three methods for generating null data. The first two work within one dyad. The third needs a sample of dyads. Choosing the right one depends on the nature of your signal and on the null hypothesis you want to test.

### 1.1 Circular Shift (`generate_surrogate_circular`)
The circular shift method takes a time series and shifts it by a random interval, wrapping the end of the series back to the beginning.

* **How it works:** It destroys the temporal alignment between Person A and Person B while perfectly preserving the exact amplitude distribution and the power spectrum of the original signal.
* **When to use it:** This is our default recommendation for most dyadic interaction data. It is computationally efficient and effectively breaks the cross-signal dependency while keeping the "rhythm" of the individual intact.
* **Important Note:** If your data has strong global trends (e.g., a person slowly moving across the room over 60 seconds), a simple circular shift might not break that trend. Ensure you use the `lag_max` argument to ensure shifts are large enough to decouple the signals.

### 1.2 Phase Randomization (`generate_surrogate_phase`)
Phase randomization uses the Fourier Transform to decompose a signal into its frequency components, randomizes the phases, and then reconstructs the signal.

* **How it works:** It creates a signal that has the exact same power spectrum (frequency content) as the original, but a totally randomized temporal structure.
* **When to use it:** This is the "gold standard" for highly oscillatory data (e.g., rhythmic postural sway, physiological signals like heart rate or respiration).
* **Important Note:** Because this method creates an entirely new signal from frequency components, it is statistically more conservative than circular shifting. It is significantly more computationally expensive to generate than circular shifts.

### 1.3 Pseudo-Dyads (`generate_surrogate_pseudo`, `generate_pseudo_dyads`)
A pseudo-dyad pairs one person's series with a partner from a *different* dyad: two people who never interacted. This is the pseudo-synchrony approach of the rMEA package (Kleinbub & Ramseyer, 2020).

* **How it works:** The partner series is real data, recorded under the same task timeline. Series of unequal length are cropped start-aligned, so sample `t` of both series sits at the same task time. `generate_pseudo_dyads()` crops each pair to its shorter series. `generate_surrogate_pseudo()` crops longer partners to the length of the target's `y`, and excludes shorter partners with a warning. Start alignment assumes that every dyad was sampled at the same rate and that sample 1 of every series is the same task onset. Resample and trim first if that is not true.
* **Which null it tests:** Circular shifts and phase randomization break all time alignment. If the task itself makes both people move at the same moments (a trial starts, a video plays), that shared movement looks like synchrony against those nulls. A pseudo-dyad partner went through the same task, so the pseudo-dyad null keeps task-driven co-movement. A significant result then means "more synchronous than with a stranger in the same setting."
* **When to use it:** When you have a sample of dyads that did the same task, and you want to separate interaction-specific synchrony from synchrony that the task produces.
* **Important Note:** A sample of N dyads gives each target at most N - 1 partners (2(N - 1) with `keep_roles = FALSE`). The smallest p-value of one dyad is 1 / (n + 1) for n partners (Section 2 gives the formula). With `keep_roles = TRUE` it is therefore never below 1 / N, so `p <= .05` needs at least 20 dyads. With `keep_roles = FALSE` the floor is 1 / (2N - 1), so `p <= .05` needs at least 11 dyads. The sample-level comparison in Section 3 uses all N(N - 1) pseudo-dyads.

## 2. Theoretical Pipeline

Regardless of the generation method, the logic of the **bsync** surrogate pipeline remains consistent:

1. **Decouple:** Generate a matrix of $N$ null time series that remove the synchrony while keeping the individual characteristics (or, for pseudo-dyads, that replace the partner with a person from another dyad).
2. **Evaluate:** Run your chosen synchrony metric (WCC, WDTW, or WGC) against every column in that matrix.
3. **Compare:** Calculate the empirical p-value as $(b + 1) / (n + 1)$, where $n$ is the number of surrogates and $b$ counts the surrogate results that are as extreme as, or more extreme than, your observed result (Phipson & Smyth, 2010). The $+ 1$ counts the observed result as one more draw from the null. So the p-value is never 0, and its smallest value is $1 / (n + 1)$. When the observed result is one more draw from the same null as the surrogates, the test rejects no more often than its nominal rate.

```{r example, eval = FALSE}
library(bsync)

# 1. Generate 1000 surrogates using your preferred method
null_matrix <- generate_surrogate_circular(
  y = dyad_data$person_B,
  n_surrogates = 1000
)

# 2. Compare observed results to the null distribution
results <- wcc_surrogate(
  x = dyad_data$person_A,
  y = dyad_data$person_B,
  y_surrogates = null_matrix,
  window_size = 90,
  lag_max = 45
)

# 3. Interpret the empirical p-value
print(results)
```

## 3. Pseudo-Dyads: A Worked Example

The pseudo-dyad generators read a *list of dyads*. Each element is a data frame (first two columns are `x` and `y`) or a list with elements `x` and `y`, the same form that `autotune_wcc()` reads. Here we simulate a sample of 10 dyads that did the same task. Every person follows one shared task timeline, plus independent noise. The dyads have no interaction-specific coupling at all.

```{r pseudo-sim}
library(bsync)
set.seed(2026)

n_dyads <- 10
n <- 1200
smooth <- function(v, k) as.numeric(stats::filter(v, rep(1 / k, k), circular = TRUE))

# One task timeline that every participant follows
task <- as.numeric(scale(smooth(cumsum(rnorm(n)), 40)))

task_only <- lapply(seq_len(n_dyads), function(i) {
  list(x = task + rnorm(n), y = task + rnorm(n))
})
```

We test each dyad twice: against circular-shift surrogates and against pseudo-dyad surrogates. `generate_surrogate_pseudo()` returns a matrix whose columns are the `y` series of the other nine dyads, ready for any surrogate wrapper.

```{r pseudo-per-dyad}
test_dyad <- function(i, y_surrogates) {
  d <- task_only[[i]]
  wcc_surrogate(
    x = d$x, y = d$y, y_surrogates = y_surrogates,
    window_size = 100, lag_max = 20, window_increment = 25
  )$p_value
}

p_circular <- vapply(seq_len(n_dyads), function(i) {
  test_dyad(i, generate_surrogate_circular(task_only[[i]]$y, 99, lag_max = 20))
}, numeric(1))

p_pseudo <- vapply(seq_len(n_dyads), function(i) {
  test_dyad(i, generate_surrogate_pseudo(task_only, dyad = i))
}, numeric(1))

# Dyads called significant (p <= .05) by the circular-shift null, out of 10
sum(p_circular <= 0.05)

# Pseudo-dyad p-values, smallest first
sort(p_pseudo)
```

The circular-shift null breaks the shared task timeline, so it reads task-driven co-movement as synchrony: it calls `r sum(p_circular <= 0.05)` of the 10 dyads significant. The pseudo-dyad null keeps that timeline. Each pseudo-dyad p-value is $(b + 1) / (9 + 1)$, where $b$ counts the strangers who score at least as high as the real partner. So the p-value is the real partner's rank among the 10 series divided by 10, and it can only be 0.1, 0.2, and so on up to 1. With 10 dyads, no single-dyad pseudo-dyad test can reach `p <= .05`. If the real partner is no different from a stranger, each rank has chance 1/10, so the test keeps its nominal size. With N dyads the smallest single-dyad p-value is 1/N, so `p <= .05` needs at least 20 dyads. With fewer dyads, use the sample-level comparison below.

For a sample-level comparison, `generate_pseudo_dyads()` builds every role-keeping pairing (the `x` of one dyad with the `y` of another: 10 × 9 = 90 pseudo-dyads). Compute the same statistic on the real dyads and on the pseudo-dyads, then compare the two distributions. Here we add a real coupling to each dyad: person B follows person A by 10 samples.

```{r pseudo-sample}
coupled <- lapply(seq_len(n_dyads), function(i) {
  own <- smooth(rnorm(n), 5)
  list(
    x = task + own + rnorm(n, sd = 0.5),
    y = task + c(rep(0, 10), own[1:(n - 10)]) + rnorm(n, sd = 0.5)
  )
})
pseudo <- generate_pseudo_dyads(coupled)
length(pseudo)
head(attr(pseudo, "sources"))

mean_abs_z <- function(d) {
  wcc(d$x, d$y, window_size = 100, lag_max = 20, window_increment = 25)$aggregate[[1]]
}
real_z <- vapply(coupled, mean_abs_z, numeric(1))
pseudo_z <- vapply(pseudo, mean_abs_z, numeric(1))

summary(real_z)
summary(pseudo_z)
```

The real dyads score higher than the pseudo-dyads, because only real partners share the 10-sample coupling. Both groups share the task timeline. How you compare the two distributions (a t-test, a permutation test, or a mixed model) is your choice. The generators only build the pseudo-dyads.

## 4. Best Practices

* **Permutation Count:** For a stable p-value, we recommend at least 1,000 permutations. Using fewer (e.g., 100) is fine for rapid exploratory analysis, but the smallest possible p-value is $1 / (n + 1)$: about .0099 with 100 surrogates, and about .001 with 1,000. A result counts as significant at $p \le .05$, so below 19 surrogates no result can be significant.
* **Power and Speed:** Surrogate analysis is "embarrassingly parallel." For heavy analyses like WDTW, always use the `future` package to distribute these 1,000+ calculations across your machine's CPU cores.
* **Interpretation:** A significant p-value does not mean the behavior is *causal*; it only means the observed degree of synchrony is statistically unlikely to have occurred if the two participants were moving independently. For a pseudo-dyad null, it means more synchrony than with a stranger in the same setting.

## References

Kleinbub, J. R., & Ramseyer, F. T. (2020). rMEA: An R package to assess nonverbal
synchronization in motion energy analysis time-series. *Psychotherapy Research*.
https://doi.org/10.1080/10503307.2020.1844334

Phipson, B., & Smyth, G. K. (2010). Permutation p-values should never be zero:
calculating exact p-values when permutations are randomly drawn. *Statistical
Applications in Genetics and Molecular Biology*, 9(1), Article 39.
https://doi.org/10.2202/1544-6115.1585
