---
title: "WCC Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WCC Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Selecting hyperparameters is often the most challenging aspect of running a Windowed Cross-Correlation (WCC) analysis. Because behavioral time series vary wildly in their speed and structure, there is no universal set of defaults. A mother-infant gaze interaction unfolds on a completely different timescale than a high-speed athletic exchange.

The `bsync` package provides the `suggest_wcc_params()` function to help you ground your parameter selection in the physical properties of your data and the theoretical boundaries of the behavior you are studying.

## Deriving Parameters from the Measured Signal

The new interface to `suggest_wcc_params()` takes the actual data (`x` and `y`) so it can
estimate the dominant behavioral cycle from the signal's own power spectral density (PSD).
This is a stronger starting point than a theoretical guess: you let the signal tell you what
its characteristic timescale is.

We illustrate with the built-in `sim_dyad` dataset (3D positional data, 80 Hz, 30 seconds).
The synchronized signals are `z_A` and `z_B`, which share a 0.5 Hz sine wave with a shifting
lead–lag relationship.

```{r example-setup}
library(bsync)

fs <- 80 # sampling rate of sim_dyad (Hz)

params <- suggest_wcc_params(
  x           = sim_dyad$z_A,
  y           = sim_dyad$z_B,
  sample_rate = fs
)
```

The PSD identifies the dominant cycle (~2 s for a 0.5 Hz signal). The 4-cycles-per-window
heuristic (Boker et al., 2002) then yields `window_size ≈ 4 × 2 s × 80 Hz = 640 samples`,
which is capped when the series is too short to fit two non-overlapping windows.

### Expert override: `event_duration_sec`

If you already know the behavioral timescale from theory or pilot data, pass it directly.
This bypasses the PSD step while still enforcing the hard constraints:

```{r example-override, eval=FALSE}
# Override: assume a 1-second event at 80 Hz
params_expert <- suggest_wcc_params(
  x                  = sim_dyad$z_A,
  y                  = sim_dyad$z_B,
  sample_rate        = fs,
  event_duration_sec = 1
)
```

### Hard constraints always applied

Regardless of how the window size is derived, `suggest_wcc_params()` enforces and reports:

- **`lag_max ≤ floor(window_size / 2)`** (SUSY reliability constraint: `segment ≥ 2·maxlag`).
  If your `max_delay_sec` would exceed this, `lag_max` is capped with a warning.
- **`window_size ≤ floor(series_length / 2)`** — at least two non-overlapping windows must
  fit in the series. A warning is issued when the derived window is too large.
- **`window_size ≥ min_window_samples`** — a minimum-samples floor for stable correlations
  (default 20). A warning is issued if the derived window falls below this.

### Window overlap (`overlap_pct`)

The `overlap_pct` parameter controls how far the window shifts on each step.

- **0% overlap:** window jumps completely forward — fastest but lowest temporal resolution.
- **50–75% overlap:** a balanced choice for exploratory analysis.
- **Increment of 1:** highest resolution; feasible for typical datasets because `bsync` uses
  a C++ prefix-sum core.

## Putting it into Practice

Once you have reviewed the suggested parameters, plug them directly into `wcc()`:

```{r example-run, eval=FALSE}
results <- wcc(
  x                = sim_dyad$z_A,
  y                = sim_dyad$z_B,
  window_size      = params$window_size,
  lag_max          = params$lag_max,
  window_increment = params$window_increment,
  lag_increment    = params$lag_increment
)
```
