---
title: "Windowed Dynamic Time Warping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Windowed Dynamic Time Warping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

While Windowed Cross-Correlation (WCC) is excellent for capturing linear relationships, it forces the data into a rigid temporal grid. If two participants are performing the same behavior but at slightly different speeds, standard correlation methods will fail to recognize the synchrony.

Windowed Dynamic Time Warping (WDTW) solves this problem. It excels at finding similarities in signals that might be distorted, stretched, or compressed in time.

It is important to remember that WDTW calculates a distance metric rather than a correlation. This means that lower values indicate stronger synchronization (a smaller distance between the signals).

## 1. What is Windowed Dynamic Time Warping?

Dynamic Time Warping finds the optimal alignment between two time series by non-linearly warping the time axis. It creates a local cost matrix and finds the lowest-cost path to map the points of Person A's signal onto the points of Person B's signal, allowing for one-to-many and many-to-one matches.

By calculating DTW over a rolling window, **bsync** allows you to track how this shape-matching distance changes over the course of an interaction.

### 1.1 Assumptions and Applicability

**When is WDTW useful?**

* **Variable Execution Speeds:** It is ideal for complex behavioral mimicry where participants perform matching actions at different paces (e.g., one person smiles quickly while the other produces a slow, drawn-out smile).
* **Shape over Timing:** When the structural pattern of the behavior is more theoretically important than the rigid frame-by-frame timing.

**When should you avoid WDTW?**

* **When Exact Duration Matters:** Because WDTW achieves alignment by stretching and compressing the time axis, it inherently distorts the temporal reality of the interaction. If you need to know exactly how many milliseconds separate two behaviors, WCC is a safer choice.
* **Highly Noisy Data:** WDTW is highly sensitive to extreme outliers and high-frequency noise. If data is not properly smoothed, the algorithm may aggressively warp the time axis to match random noise spikes rather than the underlying behavioral signal.
* **Strict Causal Inference:** Like WCC, WDTW measures alignment and similarity. It cannot statistically prove that one signal is causing the other, which requires predictive methods like Windowed Granger Causality.

## 2. Simulating Realistic Dyadic Data

To demonstrate the workflow, we will simulate an interaction between two participants (Person A and Person B) captured at 30 Hz. We will generate smooth continuous data to simulate bodily motion.

In this scenario, Person A leads the interaction by 15 frames (0.5 seconds) at the start. Over the course of the 60 seconds, the dynamic smoothly transitions until Person B leads by 15 frames at the end.

```{r setup-data}
library(bsync)

set.seed(2026)

# Simulation parameters
fs <- 30
n_frames <- 1800 # 60 seconds of data

# Generate a smooth base signal using a 30-frame moving average.
raw_noise <- rnorm(n_frames + 300)
base_signal <- stats::filter(raw_noise, rep(1 / 30, 30), circular = TRUE)
base_signal <- as.numeric(base_signal)

person_A <- numeric(n_frames)
person_B <- numeric(n_frames)

# Continuous lag shift: Person A leads by 15 frames, smoothly transitioning to B leading
lag_shifts <- round(seq(15, -15, length.out = n_frames))

for (i in 1:n_frames) {
  idx_A <- 150 + i
  idx_B <- 150 + i - lag_shifts[i]

  # Add slight independent noise to mimic realistic measurement error
  person_A[i] <- base_signal[idx_A] + rnorm(1, sd = 0.05)
  person_B[i] <- base_signal[idx_B] + rnorm(1, sd = 0.05)
}

dyad_data <- data.frame(
  time = seq(0, by = 1 / fs, length.out = n_frames),
  person_A = person_A,
  person_B = person_B
)
```

## 3. Calculating Windowed Dynamic Time Warping

With our data ready, we can run the primary `wdtw()` function. This function slides a window across the time series and calculates the DTW alignment distance at various lags within each window.

We strongly recommend setting `scale_method` to either `"global"` or `"local"`. DTW distances are highly sensitive to the scale of the input variables. Standardizing the data ensures that the resulting cost matrix is driven by the structural shape of the behaviors rather than arbitrary measurement units. For highly non-stationary data where baselines drift significantly over time, consider using `"local"`.

The `distance_metric` defaults to `"L2"` (squared Euclidean distance), which is standard for DTW, but `"L1"` (Manhattan) is also available.

```{r calculate-wdtw}
wdtw_results <- wdtw(
  x = dyad_data$person_A,
  y = dyad_data$person_B,
  time = dyad_data$time,
  window_size = 90,
  lag_max = 45,
  window_increment = 30,
  lag_increment = 1,
  scale_method = "global",
  distance_metric = "L2"
)

# View a summary of the results
print(wdtw_results)
```

The `wdtw()` function returns a list object of class `wdtw_res` containing the results data frame, the overall mean distance, and the input settings.

## 4. Surrogate Testing for Significance

Because DTW is an alignment algorithm, it will almost always find some path to align two signals, resulting in a distance cost. To determine if this distance is significantly smaller (i.e., better aligned) than what would be expected by chance, we use surrogate testing.

First, we generate a null matrix using phase randomization, which is highly recommended for continuous, oscillatory data. Then, we pass it to `wdtw_surrogate()`.

**Managing the "DTW Wall":**
Windowed Dynamic Time Warping is notoriously computationally heavy because it calculates a full cost matrix for every possible alignment in every window. Running 1,000 permutations sequentially can take a very long time.

To handle this, **bsync** offers two solutions:
1. **Parallel Processing:** By loading the `future` package and setting a `plan()`, **bsync** automatically distributes the 1,000 permutations across your available CPU cores.
2. **The Fast Method:** For exploratory analysis, setting `fast_method = TRUE` evaluates the surrogates only at a lag of 0. This reduces computation time by roughly 99 percent while still providing a reasonable estimate of chance alignment. For final publications, you should always run the full grid (`fast_method = FALSE`).

```{r wdtw-surrogate, eval=FALSE}
# 1. Distribute the workload across 4 CPU cores
library(future)
plan(multisession, workers = 4)

# 2. Generate null data preserving the amplitude and power spectrum
surrogate_matrix <- generate_surrogate_phase(
  y = dyad_data$person_B,
  n_surrogates = 1000
)

# 3. Evaluate the observed distance against the null distances
surrogate_results <- wdtw_surrogate(
  x = dyad_data$person_A,
  y = dyad_data$person_B,
  y_surrogates = surrogate_matrix,
  time = dyad_data$time,
  window_size = 90,
  lag_max = 45,
  window_increment = 30,
  lag_increment = 1,
  scale_method = "global",
  distance_metric = "L2",
  fast_method = FALSE
)

plan(sequential)

print(surrogate_results)
#> ── WDTW Surrogate Analysis (Pseudo-Synchrony) ──────────────────────────────────────────────────────────────────
#> Permutations: 1000
#> Observed Mean Cost: 27.1612
#> Average Null Cost: 57.2509
#> Empirical p-value: < 0.001
#> ✔ Observed cost is significantly lower than chance (stronger alignment).
```

The output gives us an empirical p-value by calculating the proportion of surrogate distance costs that are less than or equal to our observed distance cost. Because WDTW calculates a distance metric rather than a correlation, lower values indicate stronger synchronization. Our observed cost (27.16) was less than half the average chance cost (57.25) and was lower than all 1,000 phase-randomized permutations. Therefore, the empirical p-value is reported as < 0.001. This statistically confirms that the shape matching we observed between Person A and Person B is driven by genuine interactive behavior.

Notice that for WDTW, a smaller cost indicates stronger synchrony. The print method automatically adjusts its logic to test if your observed cost is significantly lower than the null distribution.

## 5. Optima Extraction

While the full distance matrix is informative, we often want to extract the precise lags where the alignment is optimal within each time window. For correlation (WCC), optimal alignment means the highest value (peaks). For distance metrics (WDTW), optimal alignment means the lowest cost (valleys).

The `pick_optima()` function handles this seamlessly. Because we are providing a `wdtw_res` object, the function automatically infers that it should execute a global search across the window to find the absolute minimum distance (`search_method = "global"` and `find_min = TRUE`).

```{r optima-extraction}
# Extract optimal alignment lags (valleys) using the default global search
wdtw_optima_df <- pick_optima(wdtw_results)

summary(wdtw_optima_df)
```

This returns a `wdtw_optima` data frame containing the elapsed time indices, the optimal lags, and the corresponding DTW distance values. The console output confirms that the search mode was successfully set to "Valleys (Minima)" using a "global" search method.

## 6. Visualizing the Results

We can visualize the shifting synchronization landscape. The `plot_optima_overlay()` function detects that it is working with a `wdtw_res` object and automatically applies a sequential viridis color palette to map the distances, while plotting the global minima points on top.

By passing the `time_step` argument, the axes are automatically converted from raw frame indices to seconds.

```{r visualization, fig.width=7, fig.height=5}
plot_optima_overlay(
  surface_obj = wdtw_results,
  optima_df = wdtw_optima_df,
  time_step = 1 / fs,
  show_zero_lag = TRUE
)
```

In the resulting plot, lighter colors represent smaller distances (stronger synchrony). You will see a diagonal track corresponding to the simulated shift in the dyad's interaction. The overlaid points map exactly to the lowest alignment costs, smoothly tracing the transition from Person A leading to Person B leading.

## 7. Quantifying Leadership Dynamics

Visualizing the optima helps to understand the general pattern, but researchers ultimately need a continuous, quantifiable metric of who is driving the interaction. The `leadership_asymmetry()` function converts the extracted optima into a bounded Leadership Asymmetry Index (LAI) ranging from -1 (y entirely leads) to 1 (x entirely leads).

Because **bsync** is designed around a consistent class structure, you can seamlessly chain the entire analytical process together using the native R pipe (`|>`):

```{r pipeline-lai, fig.width=7, fig.height=5}
# Run the complete pipeline from results to LAI visualization
wdtw_results |>
  pick_optima() |>
  leadership_asymmetry(epoch_size = 10, min_valid = 3) |>
  plot(smooth = TRUE)
```

By grouping the windows into local epochs, the LAI function smooths over momentary frame-by-frame jitter to reveal the broader structural periods of dominance. The resulting plot makes the continuous transition of leadership from Person A to Person B perfectly clear.
