---
title: "Windowed Phase Synchrony"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Windowed Phase Synchrony}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a complete windowed phase synchrony analysis using the **bsync** package.

Windowed Cross-Correlation (WCC) asks whether two signals rise and fall together. Phase synchrony asks a narrower question: do two rhythmic signals keep a stable timing relationship within each cycle, whatever their amplitudes? Two people swaying, rocking, or nodding at a shared tempo can be strongly phase-locked even when one moves with much larger amplitude than the other.

## 1. What is Windowed Phase Synchrony?

`wphase()` extracts the instantaneous phase of each series once, with the analytic signal (the Hilbert transform). For each window and lag, it then summarizes the phase differences between the two series in two numbers:

* **PLV (phase-locking value):** the length of the mean unit vector of the phase differences, from 0 to 1. A PLV of 1 means the phase difference stays constant across the window. A PLV near 0 means the phase difference is spread evenly around the circle. This is the phase-locking value of Lachaux et al. (1999), averaged over the time samples of a window instead of over trials.
* **Relative phase (`rel_phase`):** the circular mean of the phase differences, in radians. Positive values mean that `x` is ahead of `y` in its cycle.

### 1.1 The Narrowband Assumption

Instantaneous phase is only well defined for a **narrowband** signal, which is a signal dominated by one rhythm. Behavioral data rarely look like that. They usually carry slow drift (posture changes, sensor drift) and fast measurement noise on top of the rhythm of interest.

This matters because the error is in the dangerous direction. Slow, broadband drift makes the phases of two series change slowly, so their difference stays nearly constant across a short window, and the PLV rises. High PLV on unfiltered data is therefore not evidence of synchrony. Lachaux et al. (1999) band-pass filter their signals before extracting the phase, and you should do the same. `wphase()` does not filter for you, because a filter changes the data and the choice of band belongs to you.

**When is windowed phase synchrony useful?**

* **Rhythmic behavior:** rocking, swaying, walking, tapping, nodding, breathing, or any behavior with a clear tempo.
* **Timing over amplitude:** when the timing relationship within each cycle matters more than how large the movements are.

**When should you avoid it?**

* **Non-rhythmic behavior:** if no single rhythm dominates the signal, phase has no clear meaning, even after filtering.
* **Missing data:** the Hilbert transform uses the whole series, so a single `NA` corrupts every phase estimate. `wphase()` stops on `NA` input. Impute with `impute_ts_gaps()` or trim with `trim_edges()` first.

## 2. Simulating Realistic Dyadic Data

We simulate two people captured at 10 Hz for 2 minutes. Both move with a shared rhythm near 0.5 Hz (one cycle every 2 seconds). The tempo wanders slowly over time, as human tempo does, and Person B follows Person A's rhythm. Person A leads by 0.5 seconds at the start, and the lead shifts smoothly until Person B leads by 0.5 seconds at the end.

On top of the rhythm, each person has independent slow drift and measurement noise.

```{r setup-data}
library(bsync)

set.seed(2026)

fs <- 10 # sampling rate in Hz
n <- 1200 # 2 minutes of data
t <- (0:(n - 1)) / fs

# A shared tempo near 0.5 Hz that wanders slowly over time
wander <- stats::filter(rnorm(n + 200, sd = 0.6), rep(1 / 50, 50), circular = TRUE)
freq <- 0.5 + as.numeric(wander)[101:(100 + n)]
phase_A <- 2 * pi * cumsum(freq) / fs

# Person B follows Person A: A leads by 0.5 s at the start, B by 0.5 s at the end
lead_sec <- seq(0.5, -0.5, length.out = n)
phase_B <- approx(t, phase_A, xout = t - lead_sec, rule = 2)$y

# Independent slow drift (a strongly autocorrelated AR(1) process) for each person
drift <- function() as.numeric(stats::arima.sim(list(ar = 0.98), n))

person_A <- cos(phase_A) + 0.25 * drift() + rnorm(n, sd = 0.3)
person_B <- cos(phase_B) + 0.25 * drift() + rnorm(n, sd = 0.3)
```

## 3. Band-Pass Filtering Before Phase Extraction

A band-pass filter keeps one frequency band and removes the rest. Here we keep 0.3 to 0.7 Hz, a band around the shared rhythm. In your own data, choose the band from theory about the behavior or from the power spectrum (see `vignette("determine-downsampling")` for `evaluate_signal_power()`).

We use a second-order Butterworth filter from the **gsignal** package and apply it with `filtfilt()`, which runs the filter forward and then backward. This zero-phase filtering does not shift the timing of the signal, which matters when timing is the thing you measure.

```{r band-pass}
bp_filter <- gsignal::butter(2, c(0.3, 0.7) / (fs / 2), type = "pass")
band_pass <- function(v) as.numeric(gsignal::filtfilt(bp_filter, v))

person_A_f <- band_pass(person_A)
person_B_f <- band_pass(person_B)
```

To see why this step matters, compare two **independent** drift series, which share nothing, before and after filtering:

```{r band-pass-demo}
indep_1 <- drift()
indep_2 <- drift()

plv_indep_raw <- wphase(indep_1, indep_2, window_size = 150, lag_max = 10)$aggregate[[1]]
plv_indep_filt <- wphase(
  band_pass(indep_1), band_pass(indep_2),
  window_size = 150, lag_max = 10
)$aggregate[[1]]
plv_dyad_raw <- wphase(person_A, person_B, window_size = 150, lag_max = 10)$aggregate[[1]]

round(c(
  independent_unfiltered = plv_indep_raw,
  independent_filtered = plv_indep_filt,
  dyad_unfiltered = plv_dyad_raw
), 3)
```

The two independent series reach a mean PLV of `r round(plv_indep_raw, 3)` without filtering, and the coupled dyad reaches only `r round(plv_dyad_raw, 3)`. On unfiltered data, the drift decides the PLV, not the coupling. After filtering, the independent pair drops to `r round(plv_indep_filt, 3)`. That value is still well above 0, because two narrowband signals keep similar phases over a short window by chance. This is why a surrogate test (Section 5) is needed to judge any PLV.

## 4. Calculating Windowed Phase Synchrony on the Filtered Pair

Phase synchrony needs windows long enough to hold several cycles. We use 15-second windows (`window_size = 150`, about 7 cycles of the 0.5 Hz rhythm) and lags up to 1 second (`lag_max = 10`), which is half a cycle. A window moves 0.5 seconds at a time (`window_increment = 5`).

```{r calculate-wphase}
wphase_results <- wphase(
  x = person_A_f,
  y = person_B_f,
  window_size = 150,
  lag_max = 10,
  window_increment = 5
)

print(wphase_results)
```

The `wphase()` function returns a list object of class `wphase_res`. Its `results_df` has one row per window (`i`, in samples) and lag (`tau`, in samples), with the `plv` and `rel_phase` of that cell.

The relative phase carries the lead–lag information. At lag 0, divide the relative phase (in radians) by `2 * pi * 0.5` (the rhythm frequency in radians per second) to get the lead of Person A in seconds:

```{r rel-phase}
lag0 <- wphase_results$results_df[wphase_results$results_df$tau == 0, ]
lag0$time_sec <- (lag0$i - 1) / fs
lag0$lead_A_sec <- lag0$rel_phase / (2 * pi * 0.5)

rows <- round(seq(1, nrow(lag0), length.out = 5))
round(lag0[rows, c("time_sec", "rel_phase", "lead_A_sec")], 2)
```

The estimated lead of Person A moves from `r round(lag0$lead_A_sec[1], 2)` seconds in the first window to `r round(lag0$lead_A_sec[nrow(lag0)], 2)` seconds in the last. The simulated lead moved from 0.5 to -0.5 seconds.

## 5. Surrogate Testing for Significance

Section 3 showed that unrelated narrowband signals still give a sizable PLV. A surrogate test asks whether the observed PLV is larger than the PLV of pairs whose timing relationship was destroyed.

Circular-shift surrogates (`generate_surrogate_circular()`) suit phase synchrony well. Each surrogate shifts the whole series of Person B by a random amount and wraps the end around to the start. The rhythm of Person B stays intact, but its alignment with Person A is broken.

One caution: a circular shift of a perfectly periodic signal only adds a constant to its phase, and PLV ignores constant phase offsets. Against a strictly periodic shared rhythm, this null has no power. Real behavior has a wandering tempo, as simulated here, which gives the test its power.

```{r wphase-surrogate}
y_surr <- generate_surrogate_circular(person_B_f, n_surrogates = 1000)

surr_mean <- wphase_surrogate(
  x = person_A_f,
  y = person_B_f,
  y_surrogates = y_surr,
  window_size = 150,
  lag_max = 10,
  window_increment = 5
)

print(surr_mean)
```

The empirical p-value is (b + 1) / (n + 1), where n is the number of surrogates and b counts the surrogates whose mean PLV is at least as large as the observed one. Here the observed mean PLV is `r round(surr_mean$observed_z, 3)`, the average surrogate gives `r round(mean(surr_mean$surrogate_z), 3)`, and p = `r round(surr_mean$p_value, 4)`.

## 6. The Peak Statistic

By default, `wphase()` summarizes the surface with the mean PLV over all windows and lags (`statistic = "mean_plv"`). The alternative, `statistic = "peak"`, takes the largest PLV across lags in each window and averages those per-window values. This is the best-lag convention that `wcc()` also offers as `statistic = "peak"`. It suits a dyad whose lead–lag shifts over time, because each window counts at the lag where the phases lock best.

Pass the same `statistic` to `wphase_surrogate()`, so that every surrogate is summarized in the same way as the observed data. The same surrogate matrix can be reused:

```{r peak-statistic}
surr_peak <- wphase_surrogate(
  x = person_A_f,
  y = person_B_f,
  y_surrogates = y_surr,
  window_size = 150,
  lag_max = 10,
  window_increment = 5,
  statistic = "peak"
)

print(surr_peak)
```

With the peak statistic, the observed value is `r round(surr_peak$observed_z, 3)` against a surrogate average of `r round(mean(surr_peak$surrogate_z), 3)`, and p = `r round(surr_peak$p_value, 4)`. The peak statistic is larger than the mean statistic for the surrogates as well as for the observed data, so compare each observed value only with the null of its own statistic. Choose the statistic before you look at the results.

## 7. Optima Extraction

`pick_optima()` finds the lag with the highest PLV in each window. With `search_method = "global"`, it returns one optimum per window.

```{r optima-extraction}
wphase_optima <- pick_optima(wphase_results, search_method = "global")

summary(wphase_optima)
```

For narrowband signals, PLV changes only slowly across lags, because a lag shift adds a nearly constant phase offset within a window. The best lag is then identified only through the tempo wander that the two people share, and it is a noisy estimate of the lead. Because we simulated the data, we can compare both estimates with the true lead at the center of each window:

```{r optima-vs-truth}
center <- wphase_optima$i + 75 # sample at the center of each 150-sample window
true_lead_samples <- lead_sec[center] * fs

cor_optima <- cor(wphase_optima$optimum_lag, true_lead_samples)
cor_rel_phase <- cor(lag0$lead_A_sec, lead_sec[lag0$i + 75])
round(c(optimum_lag = cor_optima, rel_phase_lead = cor_rel_phase), 2)
```

The PLV optima correlate `r round(cor_optima, 2)` with the true lead, and the lead read from the relative phase at lag 0 correlates `r round(cor_rel_phase, 2)`. For phase synchrony, read the lead–lag from the relative phase. Use the optima as a rough guide, and do not pass them to `leadership_asymmetry()` without checking them against the relative phase.

## 8. Visualizing the Results

`plot()` draws the PLV surface as a heatmap of windows by lags. The `time_step` argument converts the axes from samples to seconds.

```{r plot-surface, fig.width=7, fig.height=5}
plot(wphase_results, time_step = 1 / fs)
```

`plot_optima_overlay()` adds the per-window optima to the same surface:

```{r plot-optima, fig.width=7, fig.height=5}
plot_optima_overlay(
  surface_obj = wphase_results,
  optima_df = wphase_optima,
  time_step = 1 / fs,
  show_zero_lag = TRUE
)
```

```{r lag-profile, include = FALSE}
plv_by_window <- split(wphase_results$results_df$plv, wphase_results$results_df$i)
median_min_plv <- median(vapply(plv_by_window, min, numeric(1)))
median_max_plv <- median(vapply(plv_by_window, max, numeric(1)))
```

The heatmap is high across most lags in every window. Across windows, the median of the lowest PLV in a window is `r round(median_min_plv, 2)`, and the median of the highest is `r round(median_max_plv, 2)`. This is the flat lag profile described in Section 7.

## See also

- `vignette("surrogate-testing")` for circular-shift, phase-randomization, IAAFT, segment, and pseudo-dyad surrogates, and which null each one tests
- `vignette("choosing-parameters")` for tools to choose and check `window_size` and `lag_max`
- `vignette("determine-downsampling")` for reading the power spectrum with `evaluate_signal_power()`
- `vignette("wcc-workflow")` for the Windowed Cross-Correlation workflow, which shares the optima and leadership pipeline
- Lachaux, J.-P., Rodriguez, E., Martinerie, J., & Varela, F. J. (1999). Measuring phase synchrony in brain signals. *Human Brain Mapping*, 8(4), 194–208.
